Turbo fast no compromise gravitational wave parameter estimation

### 2023-01-21 src/tex/ms.tex
Update BNS result

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{color}
%\usepackage[dvipsnames]{xcolor}

\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{pyRed}{RGB}{214, 39, 40}
\newcommand{\te}[1]{\textbf{\color{pyRed}(TE: #1)}}
\newcommand{\mi}[1]{\textsf{\color{teal}[\textbf{MI:} #1]}}

@@

% Abstract with filler text
\begin{abstract}
We present a light-weight, flexible, and high-performance framework for
inferring the properties of gravitational-wave events. By combining likelihood
heterodyning, automatically-differentiable and accelerator-compatible
waveforms, and gradient-based Markov chain Monte Carlo (MCMC) sampling enhanced
by normalizing flows, we achieve full Bayesian parameter estimation for real
events like GW150914 and GW170817 within a minute of sampling time.  \mi{should
we say a standard analysis takes hours?} Our framework does not require
pre-training or explicit reparameterizations and can be generalized to handle
higher dimensional problems. We present the details of our implementation and
discuss trade-offs and future developments in the context of other proposed
strategies for real-time parameter estimation. Our code for generating the
manuscript and running the analysis is publicly available on GitHub.
\end{abstract}

@@

% Brief description of PE
Parameter estimation (PE) is one of the most commonly performed data analysis
tasks in gravitational-wave (GW) data analysis, and it underpins all of GW physics
and astrophysics \cite{Christensen:2022bxb,
2019PASA...36...10T}. The central goal of PE is to infer the parameters of a
particular GW source given the strain data detected by instruments like LIGO \cite{LIGOScientific:2014pky}, Virgo \cite{VIRGO:2014yos}
and KAGRA \cite{KAGRA:2020tym}.
In the standard compact binary coalescence (CBC) scenario, this
could mean inferring intrinsic parameters such the masses and spins of the
compact objects, as well as extrinsic parameters such as their sky localization and distance
from Earth. PE is also applied to test general relativity (GR) and constrain deviations away from its predictions in observed data
\cite{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}. PE
is a crucial step in GW science, since it translates characteristics of the
strain data into astrophysically relevant quantities that can be used to
constrain astrophysical phenomena, including informing theories of binary evolution
\cite{LIGOScientific:2021psn} and measuring the properties of nuclear matter.

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including \textsc{LALInference}
\cite{Veitch:2014wba}, \textsc{PyCBC Inference} \cite{Biwer:2018osg}, and \textsc{Bilby}
\cite{Ashton:2018jfp}.
These packages have been tested by a number of groups and
are well regarded as the standard tools. However, while these tools have passed
many robustness tests, they are known to be computationally intensive. The exact
amount of time needed to analyze one event depends on factors like the duration and frequency of the signal, as well as features of the specific waveform model.
Typical runtimes for production-level analyses of shorter events, such as binary
black hole (BBH) mergers, can range from hours to weeks, whereas longer events such as binary neutron star (BNS) mergers can reach months on a standard computer cluster.
\mi{do we have any reference for these numbers? this will be attacked otherwise, since the comparisons are not strictly fair---e.g., it takes me 2 days to run on GW170817 with IMRD, not months}
This expense precludes iterating quickly on results, launching large scale measurement simulations, or obtaining results in low latency to inform astronomers for potential followup in real time.

Additionally, in
the coming decade, there are planned upgrades for existing facilities, as well as
plans for next-generation detectors such as the Einstein Telescope (ET)
\cite{Punturo:2010zz} and the Cosmic Explorer (CE)
\cite{LIGOScientific:2016wof}. These upgrades will increase the sensitivity of
the instruments and allow for the detection of more events with better
signal-to-noise ratio (SNR). The number of events that will be detected in the
coming decade is expected to grow from around a thousand per year to over a
million per year \cite{Baibhav:2019gxm}. This will put a significant strain on
the current PE tools.

% Effort going to new tools development.
In order to address this, there are efforts from multiple groups developing
tools to speed up the PE process. This includes methods that employ modern
tools such as deep learning-based method pre-trained on a large collection of
waveforms \cite{Dax:2021tsq,Dax:2022pxd}, as well as methods that simplify the
computational challenge by leveraging our knowledge about GW signals
\cite{Islam:2022afg,Roulet:2022kot}. While these methods are promising avenues
to tackle standard GW analysis tasks in the coming decade, particularly about
events involving CBCs in GR, they rely on assumptions that may not be valid for
analysis involving additional physical effects such as lensing and beyond-GR
analysis, or may use approximations that do not compute the Bayesian likelihood
exactly.

@@

In this work, we present a light-weight, flexible, and high performance
framework to infer GW event parameters in a fully-Bayesian analysis. Our
framework implements the following major features to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item differentiable waveform models,
\item normalizing-flow enhanced Markov chain Monte-Carlo (MCMC) sampler,
\item heterodyned likelihood,
\item native support for hardware accelerators.
\end{enumerate}
The main advantage of our framework is it does not rely on specific models
\mi{what do you mean by ``models'' here? we rely on specific waveforms} or
assumptions about the problem to achieve its performance. This makes our method
extensible to problems beyond the standard CBC analysis, without sacrificing
accuracy for efficiency.
\mi{should we also say something of the advantage of having better, modern
sampling with diffable WFs? can we say we expect this to be more robust to
strange posterior geometries, as you might get by introducing tidal and non-GR
params?}

@@

The rest of the paper is structured as follows: we review the basics of PE and
introduce our framework in Sec.~\ref{sec: PE}; we present benchmarking
results on both simulated and real data in Sec.~\ref{sec: Result}; and, finally,
we discuss the implications of this work and directions for future development
in Sec.~\ref{sec: Discussion}.

@@

The main objective in PE is to obtain a multidimensional posterior distribution
$p(\mathbf{\theta} \mid d)$ on parameters $\mathbf{\theta}$ given strain
data $d$.  Such probability density represents our best inference of
the source properties, an encodes all relevant information contained in the
observed data.
To compute this object, we use Bayes' theorem to write
\begin{align}
    p(\theta \mid d) = \frac{\mathcal{L}(d \mid \theta)\pi(\theta)}{p(d)}\, ,
\end{align}
where $\mathcal{L}(d \mid \theta)$ is the likelihood function, $\pi(\theta)$ is the
prior distribution, and $p(d)$ is the evidence. Since the evidence is a
normalization constant that does not depend on the source parameters, it is
often omitted if we are only interested in the posterior distribution. The prior
distribution is often chosen to be some simple distribution, such as uniformly
distributed in the component masses or a Gaussian distribution in the spins, or it could encode astrophysical information.
Assuming the noise is drawn from a Gaussian process, the
log-likelihood for GW data is given by
\begin{align}
    \log{\mathcal{L}(d \mid \theta)} = -\frac{1}{2} \left\langle d-h(\mathbf{\theta}) \mid d-h(\mathbf{\theta})\right\rangle,
\label{eq: loglikelihood}
\end{align}
where $d$ is the observed strain data, $h(\mathbf{\theta})$ is the signal predicted by a
waveform model with a specific set of source parameters $\theta$. The right hand side of
Eq.~\eqref{eq: loglikelihood} can be evaluated in either the time or frequency domains.
For stationary noise, it is computationally cheaper to compute the likelihood in the frequency
domain, and the noise-weighted inner product can be written 
\begin{align}
    % \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
    \left\langle a \mid b\right\rangle = 4 \Re \int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)}\, \mathrm{d}f \, ,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD).
In practice, the integral becomes a discrete sum over a finite number of
samples determined by the sampling rate of the detector data and duration of
the observation.

To compute
the integral shown in Eq.~\eqref{eq: innerproduct}, we need to evaluate a chosen
waveform model $h(\mathbf{\theta})$ at a number of frequency sample points.
This makes evaluating the likelihood function often the most computationally intensive
part of PE. The most accurate waveform model is numerical relativity (NR),
which obtains waveforms by directly solving the Einstein equations numerically
for a given system.
However, depending on the source parameters, generating one time series of
strain can take a day to half a year, which makes NR prohibitively expensive for
PE. To circumvent this problem, there are several families of waveform ``approximants'', including the \textsc{IMRPhenom} family \cite{Khan:2015jqa,
Garcia-Quiros:2020qpx}, the \textsc{SEOB} family \cite{PhysRevD.89.061502}, and the NR surrogate
family\cite{Varma:2019csw}. For shorter events, such as a $30-30\ M_{\odot}$ BBH, one
waveform call could take $10\text{ms}$ to ${\sim}1s$ \kw{verify number in
lalsuite}. For longer events, such as a $1.4-1.4\ M_{\odot}$ BNS, the
evaluation time could go up to \kw{Fill}. Since one needs to evaluate the
likelihood millions of times during sampling,\footnote{A typical PE run with
\texttt{Bilby} takes ${>}10^6$ likelihood evaluations to converge.} the
computational cost in evaluating the waveform accumulates and is the main reason
of the long runtime of GW PE.

@@

Since the computational cost of evaluating a waveform model scales linearly
with the number of sample points either in the time or frequency domain, the
computational burden for longer-duration signals is often quite large. To
reduce the computational cost, there are a number of methods to reduce the
number of basis points one would need to compute the likelihood faithfully
\cite{Field:2011mf, Field:2013cfa, Smith:2016qas, Vinciguerra:2017ngf}. In this
work, we use likelihood heterodyning \cite{Cornish:2021lje} (also named
relative binning in \cite{Zackay:2018qdy}).

The idea behind the heterodyned likelihood can be summarized as follows: the
integrand in Eq.~\eqref{eq: innerproduct} is a highly oscillatory function, so
one has to sample the integrand with sufficiently dense sampling to compute the
integral faithfully. The number of sample points needed would be much smaller
if the integrand was smooth. Given a pair of waveform parameters
$\mathbf{\theta}$ and $\mathbf{\theta_0}$ that are close to each other, the
waveforms generated using the pair of parameters are similar to each other,
this means the ratio between the waveforms is a smoothly varying function.
Given a reference waveform $h(\mathbf{\theta_0})$, we can exploit this
similarity between waveforms to reduce the number of sample points needed to
compute the likelihood for the set of $\mathbf{\theta}$ that is similar to
$\mathbf{\theta_0}$. We decompose the integrand into two parts: (1) a highly
oscillatory part that depends only on the reference waveform given by
$\mathbf{\theta}$ and the data, and hence only needs to be evaluated once; and,
(2) a smoothly varying part that depends on the target waveform parameters
$\theta$, that needs to be evaluated for every new likelihood evaluation.
Because the part that depends on the target waveform parameters is smooth, we
can use far fewer sample points to compute the integral with sufficient
accuracy.

One may be concerned by the accuracy of this approximation over the target
parameter space, especially in the region where the generated waveform is
significantly different from the reference waveform. However, given that we are
interested in the most probable set of parameters, if we choose the reference
waveform to be close to the data, the waveforms that are different from the
reference waveform should necessarily also differ significantly from the data.
This means that the likelihood value for these waveforms should be
significantly smaller than the likelihood of the waveforms that are similar to
the reference waveform, and hence will not be relevant for the PE result. In
practice, one will first optimize the likelihood function with full frequency
resolution to obtain the reference waveform parameters, which can be run at a
much lower cost compared to PE.

We now give a concise description of the implementation of this approach in our
code; for a more extensive derivation of heterodyned likelihood, we refer the
reader to the reference \cite{Zackay:2018qdy}. In the heterodyned likelihood
framework, the two terms involving $h$ obtained by expanding Eq.~\eqref{eq: loglikelihood} can be approximated as
\begin{subequations} \label{eq: heterodynedlikelihood}
\begin{equation}
    \langle d \mid h \rangle \approx \sum_b \left[ A_0(b)\, r^*_0(h,b) + A_1(b)\, r^*_1(h,b) \right] ,
\end{equation}
\begin{align}
    \langle h \mid h \rangle \approx \sum_b &\left[ B_0(b)\, |r_0(h,b)|^2 + \right. \nonumber \\
    &\left. 2 B_1(b)\, \Re\{r_0(h,b)\, r_1(h,b)\} \right] 
\end{align}
\end{subequations} 
where $b$ denotes the index of a \textit{sparse} set of bins over which the integrand
will be computed; $A_0(b)$, $A_1(b)$, $B_0(b)$, and $B_1(b)$ are the
heterodyning coefficients computed using the data and the reference waveform;
and, finally, $r_0(h,b)$ and $r_1(h,b)$ are the ratios between the target waveform and the
reference waveform at the center of the bin \mi{mention difference between $r_{0/1}$?}.
To evaluate Eq.~\eqref{eq:
heterodynedlikelihood}, we need to first choose a binning scheme,
then evaluate the coefficients given the data and the reference waveform, and last the
ratio between the target waveform and the reference waveform at the center of
each bin.
\mi{this paragraph is kind of floating around, maybe it should go at the end of
this subsection?}

We can obtain further insight by noting that the Fourier phase of a GW inspiral
can be expressed as a power series $\Psi(f) = \sum_i \alpha_i f^{\gamma_i}$,
where $\alpha_i$ are some coefficients depending on the waveform parameters and
$\gamma_i$ are powers motivated by post-Newtonian theory \cite{}. For example,
for the term $\gamma_i = -5/3$, $\alpha_i$ is related to the chirp mass. The
maximum dephasing one can accrue over a frequency interval
$[f_{\textrm{min}},f_{\textrm{max}}]$ is thus given by
\begin{align}
    \delta \Psi_{\textrm{max}}(f) = 2\pi \sum_{i} (f/f_{*,i})^{\gamma_i} \textrm{sgn}(\gamma_i),
\label{eq: maxdephasing}
\end{align}
where $f_{*,i} = f_{\textrm{max}}$ for $\gamma_i \geq 0$ and $f_{*,i} =
f_{\textrm{min}}$ for $\gamma_i<0$. Given the relation shown in Eq.~\eqref{eq:
maxdephasing}, we can choose the binning scheme to divide the entire
frequency band of interest into a set of bins such that the maximum dephasing
within each bin is smaller than a certain threshold $\epsilon$, i.e.,
$|\delta\Psi_{\textrm{max}}(f_{\textrm{max}}) -
\delta\Psi_{\textrm{max}}(f_{\textrm{min}})| < \epsilon$.

For sufficiently fine bins, the ratio between the target waveform
and the reference within a bin can be approximated by linear interpolation,
\begin{align}
r(f) = \frac{h(f)}{h_0(f)} = r_0(h,b) + r_1(h,b)(f- f_m(b)) + \cdots,
\label{eq:definer}
\end{align}
\mi{I think this eqn should appear right after Eq.~\eqref{eq:
heterodynedlikelihood}, because it's necessary to understand what the terms in
that eqn are, and you don't need Eq.~\eqref{eq: maxdephasing} to understand this one}
where $b$ is the index of a particular bin, $r_0(h,b)$ and $r_1(h,b)$ are the
value and slope of the ratio at the center of the bin respectively, and
$f_m(b)$ is the center frequency of the bin. Since we have access to both $h(f)$
and $h_0(f)$, we can compute $r_0$ and $r_1$ by evaluating the value of $r(f)$
at the edge of the bin and inverting Eq.~\eqref{eq:definer}.

The final ingredient we need is the heterodyning coefficients for the data and
the reference waveform on the sparse bins, which are explicitly given by
\begin{align}
    A_0(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)}{S_n(f)} \Delta f, \\
    A_1(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)(f-f_m(b))}{S_n(f)} \Delta f, \\
    B_0(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2}{S_n(f)} \Delta f, \\
    B_1(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2(f-f_m(b))}{S_n(f)} \Delta f.
\end{align}
Note that the sum within each bin should be done with the same sampling rate as
the data, i.e., the same one would do without using the heterodyned likelihood.

@@

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW150914.png}
\caption{
    hi
}
\label{fig:GW150914}
\end{figure}

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW170817.png}
\caption{
    hi
}
\label{fig:GW170817}
\end{figure}

>>>>

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{color}

\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{pyRed}{RGB}{214, 39, 40}
\newcommand{\te}[1]{\textbf{\color{pyRed}(TE: #1)}}

@@

% Abstract with filler text
\begin{abstract}
We present a light-weighted, flexible, and high-performance framework to
estimate gravitational wave event parameters. By combining heterodyned
likelihood, automatically differentiable and accelerators compatible waveforms,
normalizing flow enhanced gradient-based Markov Chain Monte Carlo (MCMC), we
achieve parameter estimation for real events like GW150914 and GW170817 within a
minute of sampling time. Our framework does not require pre-training or explicit
reparameterization and can be generalized to handle higher dimensional problems.
We also discuss as near real-time parameter estimation has been shown to be
possible by multiple groups, what are the trade-offs and future developments the
community should be aware of. Our code for generating the manuscript and running
the analysis is publically available on GitHub.
\end{abstract}

@@

% Brief description of PE
Parameter estimation (PE) is one of the most commonly performed data analysis
tasks in gravitational wave (GW) data analysis. The central question of PE is to
infer the parameters of a particular GW model given the strain data
\cite{Christensen:2022bxb, 2019PASA...36...10T}. In the standard compact binary
coalescence scenario, this could mean inferring intrinsic parameters such the
masses and spins of the event, as well as extrinsic parameters such as sky
localization and distance from Earth to the event. PE is also used in testing
general relativity (GR), which the main question is then to check for deviations
in parameters that are related to non-GR modifications
\cite{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}. PE
is a crucial step in GW science, since it translate characteristics of the
strain data into astrophysically relevant quantities that can be used to
constrain astrophysical phenomenon such as theories of binary evolution
\cite{LIGOScientific:2021psn}.

% Old codes have trouble handling next gen data
The PE results in the most recent catalog of GW events are powered by several
community-developed PE codes, including \texttt{lalinference}
\cite{Veitch:2014wba}, \texttt{pycbc} \cite{Biwer:2018osg}, and \texttt{bilby}
\cite{Ashton:2018jfp}. These packages have been tested by a number of groups and
are well regarded as the standard tools. While the standard tools have passed
many robustness tests, they are known to be computationally intensive. The exact
amount of time needed to analyze one event depends on the length of the event.
These community tools take about a week to analyze shorter events such as binary
black hole (BBH) mergers, whereas they can take up to a month on a computer
cluster to analyze longer events such as binary neutron star (BNS) mergers. In
the coming decade, there are planned upgrade for existing facilities and
next-generation detector such as the Einstein Telescope (ET)
\cite{Punturo:2010zz} and the Cosmic Explorer (CE)
\cite{LIGOScientific:2016wof}. These upgrades will increase the sensitivity of
the detectors and allow for the detection of more events with better
signal-to-noise ratio (SNR). The number of events that will be detected in the
coming decade is expected to grow from around a thousand per year to over a
million per year \cite{Baibhav:2019gxm}. This will put a significant strain on
the current PE tools, which are not designed to handle this many events.

% Effort going to new tools development.
In order to handle the future catalog, there are efforts from multiple groups
developing tools to speed up the PE process. This includes methods that employ
modern tools such as deep learning-based method pre-trained on a large
collection of waveforms \cite{Dax:2021tsq,Dax:2022pxd}, as well as methods that
simplify the computational challenge by leveraging our knowledge about GW
signals \cite{Islam:2022afg,Roulet:2022kot}. While these methods are promising
avenues to tackle standard GW analysis tasks in the coming decade, particularly
about events involving compact binary coalescence (CBC) in GR, they capitalize
assumptions that may not be valid for analysis involving additional physical
effects such as lensing and beyond-GR analysis.

@@

In this work, we present a light-weighted, flexible, and high performance
framework to estimate GW event parameters. Our framework implements the
following major features to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item Differential waveform models
\item Normalizing flow enhanced MCMC sampler
\item Heterodyned likelihood
\item Native support for accelerator
\end{enumerate}
The main advantage of our framework is it does not rely on specific models or
assumptions about the problem to achieve its performance. This makes our method
extensible to problems beyond the standard CBC analysis.

@@

The rest of the paper is structured as follows: We review the basics of PE and
introduce our framework in section \ref{sec: PE}. We present benchmarking
results on both simulated and real data in section \ref{sec: Result}. Finally,
we discuss the implications of this work and directions for future development
in section \ref{sec: Discussion}.

@@

The main objective in PE can be summarized as given data in the form of a time
series of strain $\mathbf{d}$, find the distribution of the source parameters
$\mathbf{\theta}$ that is compatible with the data, i.e.
$p(\mathbf{\theta}|\mathbf{d})$. To compute this object, we can rewrite
it using Bayes' theorem,
\begin{align}
    p(\theta| d) = \frac{\mathcal{L}(d|\theta)\pi(\theta)}{p(d)},
\end{align}
where $\mathcal{L}(d|\theta)$ is the likelihood function, $\pi(\theta)$ is the
prior distribution, and $p(d)$ is the evidence. Since the evidence is a
normalization constant that does not depend on the source parameters, it is
often omitted if we are only interested in the posterior distribution. The prior
distribution is often chosen to be some simple distribution, such as uniformly
distributed in the component masses or a Gaussian distribution in the spins.
Assuming the noise follows from a stationary Gaussian process, the
log-likelihood in the case of GW is given by
\begin{align}
    \log{\mathcal{L}(d|\theta)} = -\left<d-h(\mathbf{\theta})|d-h(\mathbf{\theta})\right>/2,
\label{eq: loglikelihood}
\end{align}
where $d$ is the strain data, $h(\mathbf{\theta})$ is the strain predicted by a
waveform model with a specific set of source parameters. The right hand side of
eq. \ref{eq: loglikelihood} can be evaluated in either time or frequency domain.
Since it is computationally cheaper to compute the likelihood in the frequency
domain, we choose to compute the likelihood throughout this work, which can be defined as 
\te{Need to take the real part?}
\begin{align}
    % \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
    \left<a|b\right> = 4 \int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD). To compute
the integral shown in eq. \ref{eq: innerproduct}, we need to evaluate a chosen
waveform model $h(\mathbf{\theta})$ at a number of frequnecy sample points.
Evaluating the likelihood function is often the most computationally intensive
part of the PE. The most accurate waveform model is numerical relativity (NR).
However, depending on the source parameters, generating one time series of
strain can take a day to half a year, which makes NR prohibitively expensive for
PE. To circumvent this problem, there are several waveform "approximant"
families, including the IMRPhenom family \cite{Khan:2015jqa,
Garcia-Quiros:2020qpx}, SEOB family \cite{PhysRevD.89.061502}, and NR surrogate
family\cite{Varma:2019csw}. For shorter events such as a $30-30\ M_{\odot}$ BBH, one
waveform call could take $10\text{ms}$ to $\sim 1s$ \kw{verify number in
lalsuite}. For longer events such as a $1.4-1.4\ M_{\odot}$ BNS event, the
evaluation time could go up to \kw{Fill}. Since one needs to evaluate the
likelihood many times during sampling\footnote{A typical PE run with
\texttt{Bilby} takes $>10^6$ likelihood evaluation to converge.}, the
computational cost in evaluating the waveform accumulate and is the main reason
of the long runtime of GW PE.

@@

Since the computational cost of evaluating a waveform model scales linearly with
the number of sample points either in the time or frequency domain, the
computational burden for longer-duration signals is often quite large. To reduce
the computational cost, there are a number of methods to reduce the number of
basis points one would need to compute the likelihood faithfully
\cite{Field:2011mf, Field:2013cfa, Smith:2016qas, Vinciguerra:2017ngf}. We use
heterodyned likelihood \cite{Cornish:2021lje} (also named relative binning in
\cite{Zackay:2018qdy}) in this work. The idea behind heterodyned likelihood can
be summarized as the following: the integrand in eq. \ref{eq: innerproduct} is a
highly oscillatory function, so one has to sample the integrand with
sufficiently dense sampling to compute the integral faithfully. The number of
sample points needed would be much smaller if the integrand was smooth. Given a
pair of waveform parameters $\mathbf{\theta}$ and $\mathbf{\theta_0}$ that are
close to each other, the waveforms generated using the pair of parameters are
similar to each other, this means the ratio between the waveforms is a smoothly
varying function. Given a reference waveform $h(\mathbf{\theta_0})$, we can
exploit this similarity between waveforms to reduce the number of sample points
needed to compute the likelihood for the set of $\mathbf{\theta}$ that is
similar to $\mathbf{\theta_0}$. We decompose the integrand into two parts: a
highly oscillatory part that depends only on the reference waveform and the
data, hence it only needs to be evaluated once; and a smoothly varying part that
depends on the target waveform parameters $\theta$, that needs to be evaluated
for every new likelihood evaluation. Because the part that depends on the target
waveform parameters is smooth, we can use far fewer sample points to compute the
integral with sufficient accuracy.

One may be concerned by the accuracy of this approximation over the target
parameter space, especially in the region where the generated waveform is
significantly different from the reference waveform. However, given we are
interested in the most probable set of parameters, if we choose the reference
waveform to be close to the data, the waveforms that are different from the
reference waveform should also differ significantly from the data. This means
the likelihood value for these waveforms should be significantly smaller than
the likelihood of the waveforms that are similar to the reference waveform,
hence will not be relevant for the PE result. In practice, one will first
optimize the likelihood function with full frequency resolution to obtain the
reference waveform parameters, which can be run at a much lower cost compared to
PE.

We give a concise description of what we implement in our code here. For a more
extensive derivation of heterodyned likelihood, we refer the reader to the
reference \cite{Zackay:2018qdy}. In the heterodyned likelihood framework, the
two terms in eq \ref{eq: loglikelihood} are given by
\begin{align}
    \langle d|h \rangle &\approx \sum_b A_0(b) r^*_0(h,b) + A_1(b) r^*_1(h,b), \nonumber \\
    \langle h|h \rangle &\approx \sum_b B_0(b) |r_0(h,b)|^2 + 2 B_1(b) \Re[r_0(h,b)r_1(h,b)],
\label{eq: heterodynedlikelihood}
\end{align}

where $b$ denotes the index of a \textit{sparse} set of bins where the integrand
will be computed on, $A_0(b)$, $A_1(b)$, $B_0(b)$, and $B_1(b)$ are the
heterodyning coefficients computed using the data and the reference waveform,
and $r_0(h,b)$ and $r_1(h,b)$ are the ratio between the target waveform and the
reference waveform at the center of the bin. To evaluate eq. \ref{eq:
heterodynedlikelihood}, we need to be able to first choose a binning scheme,
then evaluate the coefficients given the data and the reference waveform and the
ratio between the target waveform and the reference waveform at the center of
the bin.

Considering the phasing of a waveform is denoted by a power series $\Psi(f) =
\sum_i \alpha_i f^{\gamma_i}$, where $\alpha_i$ are some coefficients depending
on the waveform parameters and $\gamma_i$ are powers motivated by post-Newtonian
theory. For example, for the term $\gamma_i = -5/3$, $\alpha_i$ is related to
the chirp mass. The maximum dephasing one can have within a frequency interval $[f_{\textrm{min}},f_{\textrm{max}}]$ is given by
\begin{align}
    \delta \Psi_{\textrm{max}}(f) = 2\pi \sum_{i} (f/f_{*,i})^{\gamma_i} \textrm{sgn}(\gamma_i),
\label{eq: maxdephasing}
\end{align}
where $f_{*,i} = f_{\textrm{max}}$ for $\gamma_i>=0$ and $f_{*,i} =
f_{\textrm{min}}$ for $\gamma_i<0$. Given the relation shown in eq. \ref{eq:
maxdephasing}, we can choose the binning scheme which divide the entire
frequency domain of interests into a set of bins such that the maximum dephasing
within each bin is smaller than a certain threshold $\epsilon$, i.e.
$|\delta\Psi_{\textrm{max}}(f_{\textrm{max}}) -
\delta\Psi_{\textrm{max}}(f_{\textrm{min}})| < \epsilon$.

Given a sufficiently dense binning scheme, the ratio between the target waveform
and the reference within a bin can be approximated by linear interpolation
\begin{align}
r(f) = \frac{h(f)}{h_0(f)} = r_0(h,b) + r_1(h,b)(f- f_m(b)) + \cdots,
\label{eq:definer}
\end{align}
where $b$ is the index of a particular bin, $r_0(h,b)$ and $r_1(h,b)$ are the
value and slope of the ratio at the center of the bin respectively, and
$f_m(b)$ is the center frequency of the bin. Since we have access to both $h(f)$
and $h_0(f)$, we can compute $r_0$ and $r_1$ by evaluating the value of $r(f)$
at the edge of the bin and invert eq. \ref{eq:definer}.

The final ingredient we need is the heterodyning coefficients given the data and
the reference waveform on the sparse bins, which are given by
\begin{align}
    A_0(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)}{S_n(f)} df, \\
    A_1(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)(f-f_m(b))}{S_n(f)} df, \\
    B_0(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2}{S_n(f)} df, \\
    B_1(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2(f-f_m(b))}{S_n(f)} df.
\end{align}
Note that the sum within each bin should be done with the same sampling rate as
the data, i.e. the same one would do in the case without using heterodyned
likelihood.

@@

\begin{figure}
    \script{plotEvents.py}
    \includegraphics[width=0.99\linewidth]{figures/GW150914.pdf}
    \caption{
        Posterior of GW150914 sampled using flowMC (blue) and Bilby (orange).
    }
    \label{fig:GW150914}
\end{figure}

\begin{figure}
    \script{plotEvents.py}
\includegraphics[width=0.99\linewidth]{figures/GW170817.pdf}
\caption{
    Posterior of GW170817 sampled using flowMC (blue) and Bilby (orange).
}
\label{fig:GW170817}
\end{figure}
